---
title: "A primer: doubly robust LATE estimation, from intuition to practice"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{A primer: doubly robust LATE estimation, from intuition to practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

This primer explains, with as little formalism as possible, what the
estimators in `drlate` do and when you would want them. It is a companion to
Słoczyński, Uysal, and Wooldridge (2022), *"Doubly Robust Estimation of
Local Average Treatment Effects Using Inverse Probability Weighted
Regression Adjustment"* (arXiv:2208.01300), whose method this package — like
the original Stata command of the same name — implements.

## 1. The problem: a treatment people choose

Suppose you want to know the effect of a treatment $D$ (a job training
program, military service, a 401(k) plan) on an outcome $Y$ (wages,
savings). People *choose* whether to take the treatment, and the ones who
take it differ from the ones who don't in ways you can't fully observe.
Comparing treated and untreated outcomes — even with regression controls —
confounds the effect of the treatment with the effect of being the kind of
person who takes it.

The classic escape is an **instrument** $Z$: a binary variable that shifts
the *likelihood* of treatment but has no direct channel to the outcome.
Draft lottery numbers, distance to a college, randomized *encouragement* to
participate — all push some people into treatment without otherwise
touching their outcomes.

Imbens and Angrist (1994) showed what an instrument can and cannot give
you. It cannot reveal the effect of treatment for everyone. It identifies
the effect for **compliers** — the people whose treatment status is
actually moved by the instrument. That quantity is the **local average
treatment effect**:

$$
\text{LATE} \;=\; \frac{\;\text{effect of } Z \text{ on } Y\;}
                      {\;\text{effect of } Z \text{ on } D\;}
$$

The numerator is the *intent-to-treat* effect; the denominator is the
*first stage* (the share of compliers). With no covariates this ratio is
the Wald estimator, and it is exactly what two-stage least squares
computes.

### Why covariates enter

Real instruments are rarely as clean as a lottery. Distance to college is
exogenous *given* family background and region; a draft lottery is clean,
but conditioning on covariates still buys precision. So in practice you
need the instrument to be as-good-as-random **conditional on covariates
$X$** — and then both the numerator and the denominator above become
covariate-adjusted contrasts. How exactly to adjust them is the entire
subject of the paper and this package.

## 2. The four estimators in one picture

Every estimator in `drlate` builds the same ratio; they differ only in how
each piece is adjusted for $X$. Each piece is an "effect of $Z$ on
something" — a problem formally identical to estimating an average
treatment effect of $Z$ under unconfoundedness — so the familiar
treatment-effects toolbox applies:

| `method` | How it adjusts | Needs correct… |
|---|---|---|
| `"ra"` | Regression adjustment: model $Y$ (and $D$) given $X$ in each instrument arm; average the predictions. | outcome/treatment models |
| `"ipw"` | Inverse probability weighting: model the **instrument propensity score** $p(X) = \Pr(Z=1 \mid X)$; reweight raw means by $1/p$ and $1/(1-p)$. | propensity score |
| `"aipw"` | Augmented IPW: regression adjustment plus an IPW correction term. | **either one** |
| `"ipwra"` | IPW **regression adjustment**: fit the outcome/treatment regressions *weighted* by the inverse propensity score. | **either one** |

The last two are **doubly robust (DR)**: they are consistent if the
propensity score model *or* the regression models are right — you get two
chances instead of one.

### Why the paper (and the package default) prefers IPWRA

AIPW achieves double robustness by *adding* a correction term, and that
sum can wander outside the logical range of the outcome — a predicted
probability of 1.3, a negative count. IPWRA achieves it by *reweighting a
quasi-likelihood regression*, so as long as you choose a sensible family
(logit for binary outcomes, Poisson for counts), fitted values stay in
range by construction. This is why IPWRA has the appealing small-sample
behavior documented in the paper's simulations, and why
`method = "ipwra"` is the default.

A second practical lesson from the paper concerns **normalization**. Raw
IPW weights $Z_i/\hat p(X_i)$ need not average to one in a sample; if a few
observations have small estimated propensity scores, unnormalized
estimators can be erratic. Normalizing the weights to sum to one (the
Hájek construction) fixes this and costs nothing, so `normalized = TRUE`
is the default — it matters only for `"ipw"` and `"aipw"`; IPWRA is
normalized by construction.

## 3. A worked example

`drlate_sim` is a simulated dataset built so that the truth is known: a
binary instrument `rsncode`, a binary treatment `nvstat` with two-sided
noncompliance (60% compliers), and a continuous outcome `lwage`. The true
effect for compliers is **0.5**. Crucially, the instrument is *not*
unconditionally random: the probability of `rsncode = 1` rises with `age`
and education, and those same variables shift the outcome.

```{r}
library(drlate)
data(drlate_sim)
str(drlate_sim)
```

### The naive answers fail

Comparing treated to untreated (even though this is "just a regression"):

```{r}
coef(lm(lwage ~ nvstat + age + educ, drlate_sim))["nvstat"]
```

Biased upward — always-takers are different people. The raw Wald ratio
ignores that the instrument favors older, more educated workers:

```{r}
with(drlate_sim,
     (mean(lwage[rsncode == 1]) - mean(lwage[rsncode == 0])) /
     (mean(nvstat[rsncode == 1]) - mean(nvstat[rsncode == 0])))
```

Twice the truth.

### The drlate answer

Three formulas, one per model: the **outcome** model, the **treatment**
model, and the **instrument propensity score** model.

```{r}
fit <- drlate(outcome    = lwage   ~ age + educ,
              treatment  = nvstat  ~ age + educ,
              instrument = rsncode ~ age + educ,
              data = drlate_sim)
summary(fit)
```

Read the three rows bottom-up:

* **ATE: Z on D** — the first stage: the instrument moves the treatment
  probability by about 0.60 (the complier share). If this is tiny, you
  have a weak instrument and the ratio is fragile.
* **ATE: Z on Y** — the intent-to-treat effect of the instrument on the
  outcome.
* **LATE: D on Y** — their ratio, the causal effect for compliers. The
  95% interval comfortably covers the true 0.5.

The standard errors here are not an afterthought: every estimation stage
(the propensity score, both outcome regressions, both treatment
regressions, and the ratio itself) is stacked into one moment system and a
joint sandwich variance is computed — the uncertainty from estimating the
propensity score propagates into the LATE's standard error, exactly as in
the Stata original.

### Seeing double robustness work

The DGP has the propensity score and the outcomes both depending on `age`
and `educ`. Give the estimator a *wrong* propensity score model
(intercept only) but correct regressions — then a *wrong* outcome model
but a correct propensity score:

```{r}
# Propensity score misspecified; regressions correct
coef(drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ 1,
            data = drlate_sim))[1]

# Regressions misspecified (intercept only); propensity score correct
coef(drlate(lwage ~ 1, nvstat ~ 1, rsncode ~ age + educ,
            data = drlate_sim))[1]
```

Either way the estimate stays near 0.5: one correct model is enough.
(Misspecify *both* and no estimator can save you.)

## 4. Choosing models and options

### Outcome and treatment families

The treatment and the outcome may each be continuous, binary, or a count;
choose families so fitted values respect the response's range — the heart
of the IPWRA recommendation:

```{r}
# Binary outcome: logit keeps fitted probabilities in [0, 1]
drlate(hijob ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
       data = drlate_sim, omodel = "logit")

# Positive outcome: Poisson (quasi-likelihood; no distributional claim)
coef(drlate(kwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
            data = drlate_sim, omodel = "poisson"))
```

### Instrument propensity score flavors

Beyond logit MLE, two estimators tailor the propensity score to its job
in the weights:

* `ivmodel = "cbps"` — covariate balancing (Imai and Ratkovic 2014):
  chooses the logit coefficients so the inverse-probability weights
  exactly balance the covariates between instrument arms.
* `ivmodel = "ipt"` — inverse probability tilting (Graham, Pinto, and
  Egel 2012): fits a separate tilted score per arm; the resulting weights
  are exactly normalized by construction.

```{r}
coef(drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
            data = drlate_sim, ivmodel = "ipt"))[1]
```

### LATT: the effect for treated compliers

`estimand = "latt"` reweights everything toward the treated population —
"among those who took the treatment, what did it do?" — identified under
somewhat weaker conditions on the instrument:

```{r}
coef(drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
            data = drlate_sim, estimand = "latt"))
```

### Overlap: the assumption you can check

IPW-type estimators divide by $\hat p(X)$ and $1-\hat p(X)$, so estimated
propensity scores at 0 or 1 are fatal. `drlate()` refuses to estimate if
any score falls outside `[pstolerance, 1 - pstolerance]`; set
`osample = TRUE` to get back a logical vector marking the violating
observations so you can inspect them and decide how to restrict the
sample. Weights (`weights =`) and clustered standard errors
(`cluster =`) are available everywhere.

## 5. Coming from Stata

| Stata | R |
|---|---|
| `drlate (lwage age_5) (nvstat age_5) (rsncode age_5)` | `drlate(lwage ~ age_5, nvstat ~ age_5, rsncode ~ age_5, data = d)` |
| `(y x, poisson)` | `omodel = "poisson"` |
| `(z x, ipt)` | `ivmodel = "ipt"` |
| `, latt` | `estimand = "latt"` |
| `, method(aipw) unnrm` | `method = "aipw", normalized = FALSE` |
| `[pw = w]` | `weights = "w"` |
| `, vce(cluster id)` | `cluster = "id"` |
| `i.educ` factor syntax | a `factor()` column in the formula |

Point estimates and standard errors agree with the Stata package to
numerical precision; the package's test suite enforces this against
fixtures generated by Stata on the SIPP extract used in the original help
file.

## References

* Słoczyński, T., S. D. Uysal, and J. M. Wooldridge (2022). "Doubly Robust
  Estimation of Local Average Treatment Effects Using Inverse Probability
  Weighted Regression Adjustment." arXiv:2208.01300.
* Imbens, G. W., and J. D. Angrist (1994). "Identification and Estimation
  of Local Average Treatment Effects." *Econometrica* 62(2), 467–475.
* Abadie, A. (2003). "Semiparametric Instrumental Variable Estimation of
  Treatment Response Models." *Journal of Econometrics* 113(2), 231–263.
* Graham, B. S., C. C. de Xavier Pinto, and D. Egel (2012). "Inverse
  Probability Tilting for Moment Condition Models with Missing Data."
  *Review of Economic Studies* 79(3), 1053–1079.
* Imai, K., and M. Ratkovic (2014). "Covariate Balancing Propensity
  Score." *JRSS-B* 76(1), 243–263.
