---
title: "Doubly robust estimation of the LATE and LATT with drlate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Doubly robust estimation of the LATE and LATT with drlate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.2, dpi = 110,
                      out.width = "100%")
```

## Overview

`drlate` estimates the local average treatment effect (LATE) and the local
average treatment effect on the treated (LATT) from observational data with
a binary instrument, following Słoczyński, Uysal, and Wooldridge (2022).
It is a faithful R port of the Stata package `drlate` (Statistical Software
Components S459708) by the same authors.

The estimation core supports:

* **Estimators** (`method`): inverse-probability-weighted regression
  adjustment (`"ipwra"`, the default, doubly robust), inverse probability
  weighting (`"ipw"`), augmented inverse probability weighting (`"aipw"`,
  doubly robust), and regression adjustment (`"ra"`).
* **Outcome and treatment models**: linear, logistic, or Poisson, so the
  outcome and the treatment may each be continuous, binary, or a count.
* **Instrument propensity score models** (`ivmodel`): logistic regression by
  maximum likelihood (default), covariate balancing (`"cbps"`, Imai and
  Ratkovic 2014), or inverse probability tilting (`"ipt"`, Graham, Pinto,
  and Egel 2012).
* **Normalized** (default) or unnormalized weighting for IPW and AIPW.
* Sampling weights and cluster-robust standard errors.

Beyond the Stata original, the package adds:

* **Diagnostics**: `plot()` displays of propensity-score overlap,
  covariate balance, and implied weights; `balance()` tables; first-stage
  strength on every printout.
* **Weak-instrument-robust inference**: Fieller/Anderson–Rubin confidence
  sets via `confint(method = "fieller")`.
* **Bootstrap inference**: `vcov = "bootstrap"` (cluster-aware,
  parallelizable).
* **The DR Hausman test** of unconfoundedness from the paper's Section 5
  (`dr_hausman()`), with an analytic standard error from a jointly
  stacked moment system.
* **Estimator comparison**: `drlate_compare()` with a dot-whisker plot.

All extensions are validated by Monte Carlo (scripts in `data-raw/` of
the source repository).

### Joint inference

Point estimates are computed from sequential weighted regressions. For
inference, the moment conditions of *every* estimation stage — the
instrument propensity score, the outcome regressions, the treatment
regressions, and the causal aggregates — are stacked into one just-identified
M-estimation system, and the variance is the sandwich
\(A^{-1} B A^{-\top} / n\) evaluated at the estimates. This reproduces the
Stata package's `gmm, onestep iterate(0)` construction: standard errors
account for the estimation uncertainty of each stage, including the
first-stage propensity score.

## Example

The bundled `drlate_sim` data simulates a binary instrument `rsncode`, a
binary treatment `nvstat` with two-sided noncompliance, and outcomes on
three scales. The true complier effect on `lwage` is 0.5.

```{r}
library(drlate)
data(drlate_sim)

fit <- drlate(lwage ~ age + educ,      # outcome model
              nvstat ~ age + educ,     # treatment model
              rsncode ~ age + educ,    # instrument propensity score model
              data = drlate_sim)
summary(fit)
```

The three reported quantities mirror the Stata package's output: the causal estimate
(LATE), the intent-to-treat effect of the instrument on the outcome
(numerator), and the first-stage effect of the instrument on the treatment
(denominator), with the LATE formed as their ratio.

```{r}
coef(fit)
confint(fit)
```

### Other estimators

```{r}
# AIPW with unnormalized moments
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
       data = drlate_sim, method = "aipw", normalized = FALSE)

# IPW: no covariates in the outcome/treatment equations
drlate(lwage ~ 1, nvstat ~ 1, rsncode ~ age + educ,
       data = drlate_sim, method = "ipw")

# Regression adjustment: no instrument covariates
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ 1,
       data = drlate_sim, method = "ra")
```

### LATT, other model families, and IPT

```{r}
# LATT with an inverse-probability-tilted instrument propensity score
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
       data = drlate_sim, estimand = "latt", ivmodel = "ipt")

# Poisson outcome model for the positive wage level
drlate(kwage ~ age + educ, nvstat ~ age + educ, rsncode ~ 1,
       data = drlate_sim, method = "ra", omodel = "poisson")
```

### Clustered standard errors and weights

```{r}
drlate(lwage ~ age, nvstat ~ age, rsncode ~ age, data = drlate_sim,
       cluster = drlate_sim$educ)
```

## Diagnostics

`plot()` provides the standard design checks — propensity-score overlap,
covariate balance before/after weighting (the love plot), and the implied
weight distributions; `balance()` returns the standardized mean
differences as a data frame:

```{r diag-balance}
fit <- drlate(lwage ~ age + educ, nvstat ~ age + educ,
              rsncode ~ age + educ, data = drlate_sim)
plot(fit, type = "balance")
balance(fit)
```

```{r diag-overlap}
plot(fit, type = "overlap")
```

## Inference beyond the default sandwich

Every printout reports the first-stage z (with z² ≈ F for a single binary
instrument) and flags weakness below F = 10. The package adds two
inference tools:

```{r}
# Weak-instrument-robust Fieller confidence set (may be unbounded when
# the first stage is weak -- that is the honest answer)
confint(fit, method = "fieller")

# Nonparametric bootstrap (percentile CIs; clusters resampled whole
# when `cluster` is supplied)
fit_b <- drlate(lwage ~ age + educ, nvstat ~ age + educ,
                rsncode ~ age + educ, data = drlate_sim,
                vcov = "bootstrap", boot_reps = 199, boot_seed = 1)
confint(fit_b)
```

## The DR Hausman test of unconfoundedness

Under one-sided noncompliance (nobody takes the treatment without the
instrument), the instrument-based LATT equals the unconfoundedness-based
ATT if treatment assignment is unconfounded given the covariates.
Section 5 of the paper turns this equality into a heterogeneity-robust
Hausman test, implemented here; the Stata package does not provide it:

```{r}
d_os <- drlate_sim
d_os$nvstat[d_os$rsncode == 0] <- 0L
dr_hausman(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
           data = d_os)
```

The simulated treatment is confounded by construction, and the test
rejects.

## Comparing estimators

```{r compare-plot}
cmp <- drlate_compare(lwage ~ age + educ, nvstat ~ age + educ,
                      rsncode ~ age + educ, data = drlate_sim)
cmp
plot(cmp)
```

## Replicating the Stata examples

The Stata help file's examples use a public SIPP extract. The equivalent R
calls are:

```{r, eval = FALSE}
sipp <- haven::read_dta("https://people.brandeis.edu/~tslocz/sipp.dta")
sipp <- subset(as.data.frame(sipp),
               !is.na(kwage) & !is.na(educ) & rsncode != 999)
sipp$lwage <- log(sipp$kwage)

# Stata: drlate (lwage age_5) (nvstat age_5) (rsncode age_5)
drlate(lwage ~ age_5, nvstat ~ age_5, rsncode ~ age_5, data = sipp)

# Stata: drlate (lwage age_5) (nvstat age_5) (rsncode age_5, ipt), latt
drlate(lwage ~ age_5, nvstat ~ age_5, rsncode ~ age_5, data = sipp,
       ivmodel = "ipt", estimand = "latt")
```

The package's test suite verifies numerical equivalence of estimates and
standard errors against fixtures generated by the Stata package on this
dataset (see `inst/stata/make-fixtures.do`).

## Citation

If you use drlate in your research, please cite both the R package and
the original Stata module it implements (see `citation("drlate")` for
BibTeX entries):

> Venkitasubramanian, K. (2026). drlate: Doubly Robust Estimation of the
> Local Average Treatment Effect in R. R package version 0.1.0.
> https://github.com/kvenkita/drlate

> Uysal, D., Słoczyński, T., & Wooldridge, J. M. (2026). DRLATE: Stata
> module to perform doubly robust estimation of the local average
> treatment effect (LATE) and the local average treatment effect on the
> treated (LATT). Statistical Software Components S459708, Boston College
> Department of Economics.

## References

* Słoczyński, T., S. D. Uysal, and J. M. Wooldridge (2022). "Doubly Robust
  Estimation of Local Average Treatment Effects Using Inverse Probability
  Weighted Regression Adjustment." arXiv:2208.01300.
* Graham, B. S., C. C. de Xavier Pinto, and D. Egel (2012). "Inverse
  Probability Tilting for Moment Condition Models with Missing Data."
  *Review of Economic Studies* 79(3), 1053–1079.
* Imai, K., and M. Ratkovic (2014). "Covariate Balancing Propensity Score."
  *JRSS-B* 76(1), 243–263.
