---
title: "Doubly robust estimation of the LATE and LATT with drlate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Doubly robust estimation of the LATE and LATT with drlate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`drlate` estimates the local average treatment effect (LATE) and the local
average treatment effect on the treated (LATT) from observational data with
a binary instrument, following Słoczyński, Uysal, and Wooldridge (2022).
It is a faithful R port of the Stata package `drlate` (Statistical Software
Components S459708) by the same authors.

The package supports:

* **Estimators** (`method`): inverse-probability-weighted regression
  adjustment (`"ipwra"`, the default, doubly robust), inverse probability
  weighting (`"ipw"`), augmented inverse probability weighting (`"aipw"`,
  doubly robust), and regression adjustment (`"ra"`).
* **Outcome and treatment models**: linear, logistic, or Poisson, so the
  outcome and the treatment may each be continuous, binary, or a count.
* **Instrument propensity score models** (`ivmodel`): logistic regression by
  maximum likelihood (default), covariate balancing (`"cbps"`, Imai and
  Ratkovic 2014), or inverse probability tilting (`"ipt"`, Graham, Pinto,
  and Egel 2012).
* **Normalized** (default) or unnormalized weighting for IPW and AIPW.
* Sampling weights and cluster-robust standard errors.

### Joint inference

Point estimates are computed from sequential weighted regressions. For
inference, the moment conditions of *every* estimation stage — the
instrument propensity score, the outcome regressions, the treatment
regressions, and the causal aggregates — are stacked into one just-identified
M-estimation system, and the variance is the sandwich
\(A^{-1} B A^{-\top} / n\) evaluated at the estimates. This reproduces the
Stata package's `gmm, onestep iterate(0)` construction: standard errors
account for the estimation uncertainty of each stage, including the
first-stage propensity score.

## Example

The bundled `drlate_sim` data simulates a binary instrument `rsncode`, a
binary treatment `nvstat` with two-sided noncompliance, and outcomes on
three scales. The true complier effect on `lwage` is 0.5.

```{r}
library(drlate)
data(drlate_sim)

fit <- drlate(lwage ~ age + educ,      # outcome model
              nvstat ~ age + educ,     # treatment model
              rsncode ~ age + educ,    # instrument propensity score model
              data = drlate_sim)
summary(fit)
```

The three reported quantities mirror the Stata package: the causal estimate
(LATE), the intent-to-treat effect of the instrument on the outcome
(numerator), and the first-stage effect of the instrument on the treatment
(denominator), with the LATE formed as their ratio.

```{r}
coef(fit)
confint(fit)
```

### Other estimators

```{r}
# AIPW with unnormalized moments
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
       data = drlate_sim, method = "aipw", normalized = FALSE)

# IPW: no covariates in the outcome/treatment equations
drlate(lwage ~ 1, nvstat ~ 1, rsncode ~ age + educ,
       data = drlate_sim, method = "ipw")

# Regression adjustment: no instrument covariates
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ 1,
       data = drlate_sim, method = "ra")
```

### LATT, other model families, and IPT

```{r}
# LATT with an inverse-probability-tilted instrument propensity score
drlate(lwage ~ age + educ, nvstat ~ age + educ, rsncode ~ age + educ,
       data = drlate_sim, estimand = "latt", ivmodel = "ipt")

# Poisson outcome model for the positive wage level
drlate(kwage ~ age + educ, nvstat ~ age + educ, rsncode ~ 1,
       data = drlate_sim, method = "ra", omodel = "poisson")
```

### Clustered standard errors and weights

```{r}
drlate(lwage ~ age, nvstat ~ age, rsncode ~ age, data = drlate_sim,
       cluster = drlate_sim$educ)
```

## Replicating the Stata examples

The Stata help file's examples use a public SIPP extract. The equivalent R
calls are:

```{r, eval = FALSE}
sipp <- haven::read_dta("https://people.brandeis.edu/~tslocz/sipp.dta")
sipp <- subset(as.data.frame(sipp),
               !is.na(kwage) & !is.na(educ) & rsncode != 999)
sipp$lwage <- log(sipp$kwage)

# Stata: drlate (lwage age_5) (nvstat age_5) (rsncode age_5)
drlate(lwage ~ age_5, nvstat ~ age_5, rsncode ~ age_5, data = sipp)

# Stata: drlate (lwage age_5) (nvstat age_5) (rsncode age_5, ipt), latt
drlate(lwage ~ age_5, nvstat ~ age_5, rsncode ~ age_5, data = sipp,
       ivmodel = "ipt", estimand = "latt")
```

The package's test suite verifies numerical equivalence of estimates and
standard errors against fixtures generated by the Stata package on this
dataset (see `inst/stata/make-fixtures.do`).

## References

* Słoczyński, T., S. D. Uysal, and J. M. Wooldridge (2022). "Doubly Robust
  Estimation of Local Average Treatment Effects Using Inverse Probability
  Weighted Regression Adjustment." arXiv:2208.01300.
* Graham, B. S., C. C. de Xavier Pinto, and D. Egel (2012). "Inverse
  Probability Tilting for Moment Condition Models with Missing Data."
  *Review of Economic Studies* 79(3), 1053–1079.
* Imai, K., and M. Ratkovic (2014). "Covariate Balancing Propensity Score."
  *JRSS-B* 76(1), 243–263.
